---
title: "Multiple Logistic Regression: Effect of Age on DEATH_EVENT"
output: html_document
---

# Background

## Research Question

Does age have an impact on whether or not someone dies in the follow-up period?

## Hypothesis

-   **H0:** the model is NOT significant in explaining whether someone dies (all non-intercept coefficients $= 0$).
-   **Ha:** the model IS significant in explaining whether someone dies (at least one coefficient $\neq 0$).

Logistic regression is appropriate here because `DEATH_EVENT` is binary (0/1). A linear probability model can produce predictions outside $[0, 1]$ and violates the constant-variance and normality assumptions of OLS. The logit link constrains predictions to $(0, 1)$ and the likelihood-ratio test directly tests the stated hypothesis.

# Setup

```{r}
df <- read.csv('/Users/kaisarthik/Library/CloudStorage/OneDrive-purdue.edu/25-26/STAT 512/STAT-512-Final-Project/heart_failure_clinical_records_dataset 2.csv')
```

# Fit the Multiple Logistic Regression

-   **Outcome:** `DEATH_EVENT` (1 = died during follow-up, 0 = survived)
-   **Primary predictor of interest:** `age`
-   **Controls:** all other baseline clinical variables

`time` (follow-up length) is intentionally EXCLUDED because it is a post-baseline administrative variable that is a consequence of the event itself, not a risk factor.

```{r}
age_model <- glm(
  DEATH_EVENT ~ age + anaemia + creatinine_phosphokinase + diabetes +
    ejection_fraction + high_blood_pressure + platelets +
    serum_creatinine + serum_sodium + sex + smoking,
  data = df,
  family = binomial(link = "logit")
)

summary(age_model)
```

# Global Likelihood-Ratio Test

Tests the stated hypothesis. Null model: intercept only. Full model: all predictors.

```{r}
null_model <- glm(DEATH_EVENT ~ 1, data = df, family = binomial(link = "logit"))
anova(null_model, age_model, test = "Chisq")

# McFadden pseudo-R^2
mcfadden <- 1 - (age_model$deviance / age_model$null.deviance)
cat(sprintf("McFadden pseudo-R^2: %.4f\n", mcfadden))
```

# Interpretation

## age

$\beta \approx 0.0557$, $p \approx 2.2 \times 10^{-5}$ (highly significant). $\text{OR} \approx \exp(0.0557) \approx 1.057$, 95% CI $\approx [1.030, 1.085]$.

Holding the other ten predictors fixed, each additional year of age multiplies the odds of dying during follow-up by about 1.057 — a \~5.7% increase in odds per year. Over a 10-year age gap, the odds of death are multiplied by roughly $\exp(0.557) \approx 1.75$, i.e. about 75% higher odds. This directly answers the research question: age has a strong, statistically significant impact on whether someone dies in the follow-up period.

# Hypothesis Test Conclusion

The likelihood-ratio test comparing the intercept-only null model to the full model gives $\chi^2 = 81.07$ on 11 df, $p \approx 9.2 \times 10^{-13}$. We therefore **REJECT** the null hypothesis and conclude that the model is significant in explaining whether someone dies during the follow-up period.

McFadden pseudo-$R^2 \approx 0.216$, which is considered a good fit for logistic regression (McFadden values of 0.2–0.4 typically indicate strong model performance).

# Reduced Model (Strip Non-Significant Predictors)

Seven predictors in the full model are not significant at $\alpha = 0.05$: `anaemia` ($p = 0.165$), `diabetes` ($p = 0.611$), `high_blood_pressure` ($p = 0.171$), `platelets` ($p = 0.661$), `serum_sodium` ($p = 0.090$), `sex` ($p = 0.255$), and `smoking` ($p = 0.697$). We fit a reduced model using only the four predictors that were significant at the 5% level in the full fit: `age`, `creatinine_phosphokinase` ($p = 0.042$), `ejection_fraction`, and `serum_creatinine`.

## Fit the reduced model

```{r}
reduced_model <- glm(
  DEATH_EVENT ~ age + creatinine_phosphokinase + ejection_fraction + serum_creatinine,
  data = df,
  family = binomial(link = "logit")
)

summary(reduced_model)
```

## General Linear Test (Likelihood-Ratio Test): Reduced vs Full

Tests whether the dropped predictors jointly contribute to the model fit.

-   **H0:** $\beta_{\text{anaemia}} = \beta_{\text{diabetes}} = \beta_{\text{HBP}} = \beta_{\text{platelets}} = \beta_{\text{sodium}} = \beta_{\text{sex}} = \beta_{\text{smoking}} = 0$ (reduced model is adequate)
-   **Ha:** at least one of the dropped coefficients is nonzero (the full model is preferred)

```{r}
glt <- anova(reduced_model, age_model, test = "Chisq")
print(glt)

cat("\nChi-square critical value (df = 7, alpha = 0.05):",
    round(qchisq(0.95, 7), 4), "\n")
cat("Chi-square test statistic:", round(glt$Deviance[2], 4), "\n")
cat("p-value:", signif(glt$`Pr(>Chi)`[2], 4), "\n")
```

## AIC / BIC comparison

```{r}
cat("Full model     — AIC:", round(AIC(age_model), 2),
    "  BIC:", round(BIC(age_model), 2), "\n")
cat("Reduced model  — AIC:", round(AIC(reduced_model), 2),
    "  BIC:", round(BIC(reduced_model), 2), "\n")

# McFadden pseudo-R^2 for the reduced model
mcfadden_red <- 1 - (reduced_model$deviance / reduced_model$null.deviance)
cat(sprintf("Reduced model McFadden pseudo-R^2: %.4f\n", mcfadden_red))
```

# Reduced Model: Interpretation

## LRT conclusion

The likelihood-ratio test gives $\chi^2 = 8.92$ on 7 df, $p = 0.259$, which is well below the critical value of $\chi^2_{0.05, 7} = 14.07$. We **fail to reject** H0: the seven dropped predictors taken together do not significantly improve fit over the reduced model. The simpler four-predictor model is therefore **statistically adequate**, and should be preferred on parsimony grounds.

## Information criteria

-   **AIC** drops from $318.28$ (full) to $313.20$ (reduced) — a 5-point improvement, consistent with the LRT.
-   **BIC** drops from $362.69$ (full) to $331.70$ (reduced) — a 31-point drop, which is a *large* BIC improvement and strongly favors the reduced model.
-   **McFadden pseudo-**$R^2$ decreases only slightly, from $0.216$ to $0.192$, confirming that the dropped predictors contributed very little explanatory power.

## Effect sizes in the reduced model

-   **age:** $\hat{\beta} = 0.0535$, OR $= 1.055$ per year (95% CI $[1.030, 1.081]$). Each additional year increases the odds of death by about 5.5% — essentially unchanged from the full-model estimate ($\hat{\beta} = 0.0557$).
-   **ejection_fraction:** $\hat{\beta} = -0.0704$, OR $= 0.932$ per unit (95% CI $[0.906, 0.959]$). A 5-point increase in ejection fraction multiplies the odds of death by $\exp(-0.35) \approx 0.70$ — a 30% reduction in the odds of death per 5-point EF improvement.
-   **serum_creatinine:** $\hat{\beta} = 0.671$, OR $= 1.956$ per mg/dL (95% CI $[1.423, 2.688]$). Each additional mg/dL **nearly doubles** the odds of death.
-   **creatinine_phosphokinase:** $\hat{\beta} = 1.94 \times 10^{-4}$, OR $= 1.0002$ per U/L (95% CI $[0.9999, 1.0005]$) — **no longer significant** in the reduced model ($p = 0.147$). Its apparent significance in the full model ($p = 0.042$) appears fragile once correlated noise predictors are removed.

## Sensitivity note: dropping CPK

Because CPK is no longer significant once the noise predictors are removed, an even more parsimonious three-predictor model (`age`, `ejection_fraction`, `serum_creatinine`) is a reasonable further simplification. It also passes the LRT against the full model ($\chi^2 = 11.00$ on 8 df, $p = 0.20$) and has virtually identical AIC ($313.28$) and better BIC. However, since CPK was significant at $\alpha = 0.05$ in the *full* model, we keep it here to faithfully answer the question "strip out all predictors that were non-significant in the full model, then test."

## Answer to the research question

With or without the non-significant predictors, **age remains a strong and statistically significant predictor of death during follow-up**. The OR for `age` in the reduced model is $1.055$ per year, and in the full model was $1.057$ per year — essentially identical. The reduced model gives a more parsimonious story without materially changing the answer: age has a clear, clinically meaningful, and statistically robust effect on the odds of death during the follow-up period, and this conclusion is not driven by confounding from (or collinearity with) the non-significant predictors that we dropped.

# Minimal Model (Drop CPK)

Following the sensitivity note above, we now formally drop CPK and refit. The minimal model contains only the three predictors that remain significant in the reduced fit: `age`, `ejection_fraction`, and `serum_creatinine`.

## Fit the minimal model

```{r}
minimal_model <- glm(
  DEATH_EVENT ~ age + ejection_fraction + serum_creatinine,
  data = df,
  family = binomial(link = "logit")
)

summary(minimal_model)
```

## GLT: Minimal vs Reduced (single test of CPK)

-   **H0:** $\beta_{\text{CPK}} = 0$ (CPK can be dropped from the reduced model)
-   **Ha:** $\beta_{\text{CPK}} \neq 0$ (CPK carries unique explanatory power)

```{r}
glt_cpk <- anova(minimal_model, reduced_model, test = "Chisq")
print(glt_cpk)

cat("\nChi-square critical (df = 1, alpha = 0.05):",
    round(qchisq(0.95, 1), 4), "\n")
cat("Chi-square test statistic:", round(glt_cpk$Deviance[2], 4), "\n")
cat("p-value:", signif(glt_cpk$`Pr(>Chi)`[2], 4), "\n")
```

## GLT: Minimal vs Full (all 8 dropped predictors)

-   **H0:** all 8 dropped coefficients equal zero (the 3-predictor model is adequate)
-   **Ha:** at least one of the 8 coefficients is nonzero

```{r}
glt_full <- anova(minimal_model, age_model, test = "Chisq")
print(glt_full)

cat("\nChi-square critical (df = 8, alpha = 0.05):",
    round(qchisq(0.95, 8), 4), "\n")
cat("Chi-square test statistic:", round(glt_full$Deviance[2], 4), "\n")
cat("p-value:", signif(glt_full$`Pr(>Chi)`[2], 4), "\n")
```

## AIC / BIC comparison of all three models

```{r}
comparison <- data.frame(
  AIC      = c(AIC(age_model), AIC(reduced_model), AIC(minimal_model)),
  BIC      = c(BIC(age_model), BIC(reduced_model), BIC(minimal_model)),
  Deviance = c(age_model$deviance, reduced_model$deviance, minimal_model$deviance),
  df_resid = c(age_model$df.residual, reduced_model$df.residual, minimal_model$df.residual),
  row.names = c("Full (11 predictors)", "Reduced (4 predictors)", "Minimal (3 predictors)")
)
print(round(comparison, 2))

# McFadden pseudo-R^2 for all three
null_dev <- age_model$null.deviance
cat("\nMcFadden pseudo-R^2:\n")
cat(sprintf("  Full:     %.4f\n", 1 - age_model$deviance     / null_dev))
cat(sprintf("  Reduced:  %.4f\n", 1 - reduced_model$deviance / null_dev))
cat(sprintf("  Minimal:  %.4f\n", 1 - minimal_model$deviance / null_dev))
```

# Minimal Model: Interpretation

## LRT conclusions

**Minimal vs Reduced** (direct test of CPK alone): $\chi^2 = 2.08$ on 1 df, $p = 0.149$, below the critical value of $\chi^2_{0.05, 1} = 3.84$. We **fail to reject** H0: the data are consistent with $\beta_{\text{CPK}} = 0$. CPK's effect is not distinguishable from zero once we condition on `age`, `ejection_fraction`, and `serum_creatinine`.

**Minimal vs Full** (testing all 8 dropped predictors jointly): $\chi^2 = 11.00$ on 8 df, $p = 0.202$, below the critical value of $\chi^2_{0.05, 8} = 15.51$. We **fail to reject** H0: the full set of 8 non-essential predictors does not significantly improve fit over the 3-predictor minimal model. The minimal model is **statistically adequate** relative to the full model.

## Information criteria

| Model                  | AIC    | BIC        | McFadden $R^2$ |
|------------------------|--------|------------|----------------|
| Full (11 predictors)   | 318.28 | 362.69     | 0.2160         |
| Reduced (4 predictors) | 313.20 | 331.70     | 0.1922         |
| Minimal (3 predictors) | 313.28 | **328.08** | 0.1867         |

-   **AIC** is essentially tied between the reduced and minimal models (313.20 vs 313.28) — AIC mildly prefers reduced because CPK is kept at the margin of significance.
-   **BIC** is minimized by the minimal model (328.08). BIC penalizes extra parameters more heavily than AIC, and it decisively selects the 3-predictor model.
-   **McFadden** pseudo-$R^2$ drops only marginally (0.192 → 0.187) — explanatory power is essentially preserved.

When AIC and BIC disagree, we lean on the formal hypothesis test and the parsimony principle: the LRT clears the minimal model ($p = 0.149$ for dropping CPK; $p = 0.202$ for dropping all 8 predictors jointly), so we adopt it.

## Effect sizes in the minimal model

-   **age:** $\hat{\beta} = 0.0517$, OR $= 1.053$ per year (95% CI $[1.028, 1.079]$), $p \approx 2.7 \times 10^{-5}$. Essentially identical to the full ($1.057$) and reduced ($1.055$) estimates.
-   **ejection_fraction:** $\hat{\beta} = -0.0700$, OR $= 0.932$ per unit (95% CI $[0.907, 0.959]$), $p \approx 8.7 \times 10^{-7}$. A 10-point improvement in EF multiplies the odds of death by $\exp(-0.70) \approx 0.50$ — roughly halves the odds of death.
-   **serum_creatinine:** $\hat{\beta} = 0.666$, OR $= 1.946$ per mg/dL (95% CI $[1.425, 2.659]$), $p \approx 2.9 \times 10^{-5}$. Each additional mg/dL nearly doubles the odds of death.

All three predictors are highly significant ($p < 10^{-4}$).

## Stability of the age effect across specifications

| Model                  | $\hat{\beta}_{\text{age}}$ | $p$-value            |
|------------------------|----------------------------|----------------------|
| Full (11 predictors)   | 0.0557                     | $2.2 \times 10^{-5}$ |
| Reduced (4 predictors) | 0.0535                     | $1.9 \times 10^{-5}$ |
| Minimal (3 predictors) | 0.0517                     | $2.7 \times 10^{-5}$ |

The estimated age effect is remarkably stable across specifications: the OR moves only from $1.057$ to $1.053$ as we strip away 8 predictors, and the p-value is essentially unchanged. This is strong evidence that age is an **independent, robust risk factor** for death during follow-up — its effect is not driven by confounding with, or collinearity against, any of the other measured clinical variables.

## Final answer to the research question

A parsimonious three-predictor model — `age`, `ejection_fraction`, and `serum_creatinine` — is statistically equivalent to the full 11-predictor model (LRT $p = 0.202$), achieves the **lowest BIC** among all three candidate models, and estimates an age effect nearly identical to that of the full model. We therefore conclude:

> **Age has a robust, statistically significant effect on the odds of death during the follow-up period.** The odds of death increase by approximately **5.3% per additional year of age** ($p \approx 2.7 \times 10^{-5}$), controlling for ejection fraction and serum creatinine. Equivalently, each 10-year increase in age raises the odds of death by approximately **68%** ($\exp(0.517) \approx 1.68$). This conclusion holds across every model specification we examined and is not an artifact of model choice.
