---
title: "Ejection Fraction Multiple Linear Regression"
output: html_document
date: "2026-04-21"
---

## Goal

Use multiple linear regression to determine which of the five selected predictors are significantly associated with `ejection_fraction`, following the project pipeline through step 5:

1.  Assumption screening
2.  Initial model
3.  Diagnostics
4.  Decision and final model
5.  Model quality and validation

## Setup

```{r setup, message=FALSE, warning=FALSE}
options(scipen = 999)

library(car)
library(caret)
library(lmtest)

data_path <- "/Users/kaisarthik/Library/CloudStorage/OneDrive-purdue.edu/25-26/STAT 512/STAT-512-Final-Project/heart_failure_clinical_records_dataset 2.csv"

df <- read.csv(data_path)
df$high_blood_pressure <- factor(
  df$high_blood_pressure,
  levels = c(0, 1),
  labels = c("No", "Yes")
)
```

## Step 1: Assumption Screening

The exploratory stage is used to check whether the MLR assumptions are plausible before making inference from the full model.

### Fit the full model

```{r}
full_model <- lm(
  ejection_fraction ~ high_blood_pressure + serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)

full_vif <- round(vif(full_model), 3)
full_bp <- bptest(full_model)
full_shapiro <- shapiro.test(residuals(full_model))
full_dw <- dwtest(full_model)
full_std_resid <- rstandard(full_model)
full_hat <- hatvalues(full_model)
full_cooks <- cooks.distance(full_model)

n <- nrow(df)
p <- length(coef(full_model))
outlier_rows <- as.integer(which(abs(full_std_resid) > 3))
high_leverage_rows <- as.integer(which(full_hat > (2 * p / n)))
influential_rows <- as.integer(which(full_cooks > (4 / n)))
```

### Assumption checks

```{r}
cat("Linearity:\n")
cat("Inspect the predictor-response plots and residual plots for curvature.\n\n")

cat("Independence:\n")
cat("Each row is one patient, so independence is plausible by design.\n")
cat(
  sprintf(
    "Durbin-Watson = %.4f, p-value = %.4f\n\n",
    unname(full_dw$statistic),
    full_dw$p.value
  )
)

cat("Multicollinearity:\n")
print(full_vif)
cat("\n")

cat("Constant variance:\n")
cat(
  sprintf(
    "Breusch-Pagan = %.4f on %d df, p-value = %.4f\n\n",
    unname(full_bp$statistic),
    unname(full_bp$parameter),
    full_bp$p.value
  )
)

cat("Normality:\n")
cat(
  sprintf(
    "Shapiro-Wilk W = %.4f, p-value = %.4f\n\n",
    full_shapiro$statistic,
    full_shapiro$p.value
  )
)

cat("Unusual observations:\n")
cat("Rows with |standardized residual| > 3:\n")
if (length(outlier_rows) == 0) {
  cat("None\n")
} else {
  print(outlier_rows)
}

cat(sprintf("\nHigh leverage threshold: h_i > %.4f\n", 2 * p / n))
if (length(high_leverage_rows) == 0) {
  cat("None\n")
} else {
  print(high_leverage_rows)
}

cat("\nCook's distance threshold: 4/n\n")
if (length(influential_rows) == 0) {
  cat("None\n")
} else {
  print(influential_rows)
}
```

### Screening plots

```{r}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

boxplot(
  ejection_fraction ~ high_blood_pressure,
  data = df,
  main = "EF by High Blood Pressure",
  xlab = "High Blood Pressure",
  ylab = "Ejection Fraction",
  col = c("grey85", "lightblue")
)

plot(
  df$serum_sodium, df$ejection_fraction,
  pch = 19, col = rgb(0, 0, 0, 0.45),
  xlab = "Serum Sodium", ylab = "Ejection Fraction",
  main = "EF vs Serum Sodium"
)
abline(lm(ejection_fraction ~ serum_sodium, data = df), col = "red", lwd = 2)

plot(
  df$serum_creatinine, df$ejection_fraction,
  pch = 19, col = rgb(0, 0, 0, 0.45),
  xlab = "Serum Creatinine", ylab = "Ejection Fraction",
  main = "EF vs Serum Creatinine"
)
abline(lm(ejection_fraction ~ serum_creatinine, data = df), col = "red", lwd = 2)

plot(
  df$age, df$ejection_fraction,
  pch = 19, col = rgb(0, 0, 0, 0.45),
  xlab = "Age", ylab = "Ejection Fraction",
  main = "EF vs Age"
)
abline(lm(ejection_fraction ~ age, data = df), col = "red", lwd = 2)
```

## Step 2: Initial Model

Research question: Which of the five selected predictors are significantly associated with `ejection_fraction`?

Global hypotheses for the full model:

-   `H0`: all five slope coefficients are 0.
-   `HA`: at least one slope coefficient is not 0.

### Full model

```{r}
summary(full_model)
```

### Coefficient table

```{r}
full_coef_table <- cbind(
  round(coef(summary(full_model)), 4),
  round(confint(full_model), 4)
)
print(full_coef_table)
```

## Step 3: Diagnostics

The full model diagnostics are used to determine whether any remedy or simplification is needed.

### Numerical diagnostics

```{r}
full_vif
full_bp
full_shapiro
full_dw
```

### Influence diagnostics

```{r}
cat("Cook's distance rows above 4/n:\n")
if (length(influential_rows) == 0) {
  cat("None\n")
} else {
  print(influential_rows)
}
```

### Diagnostic plots

```{r}
par(mfrow = c(2, 2))
plot(full_model)
```

## Step 4: Decision and Final Model

A minimal remedy is considered first by replacing raw `creatinine_phosphokinase` with `log_cpk`, since raw CPK is highly right-skewed. Then reduced nested models are compared to identify the most interpretable model without losing predictive quality.

### Fit candidate models

```{r}
df$log_cpk <- log(df$creatinine_phosphokinase)

log_cpk_model <- lm(
  ejection_fraction ~ high_blood_pressure + serum_sodium +
    serum_creatinine + log_cpk + age,
  data = df
)

sodium_only_model <- lm(
  ejection_fraction ~ serum_sodium,
  data = df
)

drop_cpk_test <- anova(no_cpk_model, full_model)
drop_hbp_creat_test <- anova(sodium_age_model, no_cpk_model)
drop_age_test <- anova(sodium_only_model, sodium_age_model)
```

```{r}
model_metrics <- data.frame(
  Model = c(
    "Full model (raw CPK)",
    "Full model (log CPK)",
    "Final: sodium only"
  ),
  Adjusted_R2 = c(
    summary(full_model)$adj.r.squared,
    summary(log_cpk_model)$adj.r.squared,
    summary(sodium_only_model)$adj.r.squared
  ),
  AIC = c(
    AIC(full_model),
    AIC(log_cpk_model),
    AIC(sodium_only_model)
  ),
  BIC = c(
    BIC(full_model),
    BIC(log_cpk_model),
    BIC(sodium_only_model)
  )
)

metric_cols <- c("Adjusted_R2", "AIC", "BIC")
model_metrics[metric_cols] <- lapply(model_metrics[metric_cols], round, 4)
model_metrics
```

### Nested model comparisons

```{r}
drop_cpk_test
drop_hbp_creat_test
drop_age_test
```

### Decision

Use `serum_sodium` only as the final model.

Reason:

-   it is the only consistently significant predictor
-   it has the smallest AIC and BIC among the candidate models
-   it has the smallest 10-fold cross-validated RMSE
-   it is the most interpretable final model

### Final model summary

```{r}
summary(sodium_only_model)
```

### Final model coefficient table

```{r}
final_coef_table <- cbind(
  round(coef(summary(sodium_only_model)), 4),
  round(confint(sodium_only_model), 4)
)
print(final_coef_table)
```

## Step 5: Model Quality and Validation

### 10-fold cross-validation

```{r}
set.seed(512)
cv_folds <- createFolds(df$ejection_fraction, k = 10, returnTrain = TRUE)
cv_control <- trainControl(method = "cv", number = 10, index = cv_folds)

set.seed(512)
cv_full <- train(
  ejection_fraction ~ high_blood_pressure + serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df,
  method = "lm",
  metric = "RMSE",
  trControl = cv_control
)

set.seed(512)
cv_log_cpk <- train(
  ejection_fraction ~ high_blood_pressure + serum_sodium +
    serum_creatinine + log_cpk + age,
  data = df,
  method = "lm",
  metric = "RMSE",
  trControl = cv_control
)

set.seed(512)
cv_sodium_only <- train(
  ejection_fraction ~ serum_sodium,
  data = df,
  method = "lm",
  metric = "RMSE",
  trControl = cv_control
)

cv_results <- data.frame(
  Model = c(
    "Full model (raw CPK)",
    "Full model (log CPK)",
    "Final: sodium only"
  ),
  CV_RMSE = c(
    cv_full$results$RMSE,
    cv_log_cpk$results$RMSE,
    cv_sodium_only$results$RMSE
  ),
  CV_RMSE_SD = c(
    cv_full$results$RMSESD,
    cv_log_cpk$results$RMSESD,
    cv_sodium_only$results$RMSESD
  )
)

cv_results[c("CV_RMSE", "CV_RMSE_SD")] <- lapply(
  cv_results[c("CV_RMSE", "CV_RMSE_SD")],
  round,
  4
)

cv_results
```

### Validation takeaway

The `serum_sodium`-only model has the best average 10-fold RMSE among the candidate models, so it is retained as the final model. Even so, the adjusted `R^2` values are small across all models, which means the dataset supports only weak prediction of `ejection_fraction` using these variables.
