---
title: "Ejection Fraction Multiple Linear Regression"
output:
  pdf_document: default
  word_document: default
  html_document: default
date: "2026-04-21"
editor_options: 
  markdown: 
    wrap: 72
---

## Steps

1.  Full Model
2.  Assumptions and Diagnostics
3.  Remedial Methods
4.  Initial Model and Research Questions
5.  Final Model and Research Questions

## Setup

```{r setup, message=FALSE, warning=FALSE}
options(scipen = 999)

library(car)
library(caret)
library(lmtest)
library(dplyr)
library(MASS)

data_path <- "heart_failure_clinical_records_dataset 2.csv"

df <- read.csv(data_path)
df$high_blood_pressure <- factor(
  df$high_blood_pressure,
  levels = c(0, 1),
  labels = c("No", "Yes")
)
```

\newpage

## Step 1: Fit the Full Model

```{r}
full_model <- lm(
  ejection_fraction ~ high_blood_pressure * serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)
```

\newpage

## Step 2: Diagnostics

```{r}
# Tests below show strong signs of multicollinearity (VIFs over 1000 for
# the two dummy variables associated with the blood pressure predictor), no
# heteroscedasticity (BP test p-value > 0.05), no sign of autocorrelation (DW 
# test p-value > 0.05), but another violation of assumptions is that of normal
# residuals where the Shapiro-Wilk test has a p-value ~ 0 (< alpha = 0.05).

vif(full_model)
bptest(full_model)
shapiro.test(residuals(full_model))
dwtest(full_model)

# Visual diagnostics of residuals
plot(fitted(full_model), resid(full_model),
     xlab = "Fitted", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

qqnorm(resid(full_model), main = "Normal Q-Q")
qqline(resid(full_model), lty = 2)


# Tests below focus on identifying outliers. This can be streamlined through
# observing the outliers on the influence plot. It is worth noting that no
# observation had a Cook's D over 0.5, and there are only X-outliers (no Y).

lev_h <- hatvalues(full_model)
t_stud <- rstudent(full_model)
cookD <- cooks.distance(full_model)

n <- nobs(full_model)
p <- length(coef(full_model))
n
p
lev_cut <- 2 * p / n
lev_cut_3 <- 3 * p / n
qt(1 - 0.05 / (2 * nrow(df)), df = nrow(df) - p - 1)
outliers_q3 <- data.frame(
  obs_index = 1:n,
  leverage_h = lev_h,
  stud_del_resid = t_stud,
  cooks_D = cookD
) |>
  mutate(
    flag_leverage_gt_3p_over_n = leverage_h > lev_cut_3,
    flag_cookD_gt_0p5 = cooks_D > 0.5,
    flag_abs_t_gt_5percentB = abs(stud_del_resid) > 
      qt(1 - 0.05 / (2 * nrow(df)), df = nrow(df) - p - 1)
  ) |>
  filter(
    flag_leverage_gt_3p_over_n |
    flag_cookD_gt_0p5 |
    flag_abs_t_gt_5percentB
  )

outliers_q3

plot(
  lev_h, t_stud,
  xlab = "Leverage (hii)",
  ylab = "Studentized Deleted Residuals (ti)",
  main = "Influence Plot",
  pch = 19
)

abline(h = 2, lty = 2)
abline(h = -2, lty = 2)
abline(h = qt(1 - 0.05 / (2 * nrow(df)), df = nrow(df) - p - 1), lty = 3)
abline(h = -qt(1 - 0.05 / (2 * nrow(df)), df = nrow(df) - p - 1), lty = 3)

abline(v = lev_cut, lty = 2, col = "blue")
abline(v = lev_cut_3, lty = 3, col = "red")

text(lev_h, t_stud, labels = 1:n, pos = 4, cex = 0.7)

png("avplots_all.png", width = 2400, height = 1800, res = 220)
par(mfrow = c(3, 3))
avPlots(full_model, id = FALSE)
dev.off()
```

\newpage

## Step 3: Remedial Methods

The main takeaway from Step 2 is that the normality assumption is not
met. Because p-value-based inference (t-tests, F-tests, nested ANOVAs)
relies on that assumption holding, the model must be corrected before
the research questions can be answered. To address this, a boxcox
transformation is applied.

```{r}
# Apply Box-cox (Topic 8R)
dat <- as.data.frame(df)
fit_raw <- lm(ejection_fraction ~ high_blood_pressure * serum_sodium +
                serum_creatinine + creatinine_phosphokinase + age, data = dat)
bc <- boxcox(fit_raw, lambda = seq(-1.5, 1.5, by = 0.05), plotit=TRUE)
lambda <- bc$x[which.max(bc$y)]
lambda

# Fit log-transformed response
fit_log <- lm(log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
                serum_creatinine + creatinine_phosphokinase + age, data = dat)

# Groupwise variance check
dat$log_ejection_fraction <- log(dat$ejection_fraction)
g_bc <- cut(fitted(fit_log), breaks = quantile(fitted(fit_log), probs = seq(0, 1, by = 0.25)),
            include.lowest = TRUE)
tapply(dat$log_ejection_fraction, g_bc, sd)


# Diagnostics on transformed scale
plot(fitted(fit_log), resid(fit_log),
     xlab = "Fitted (log Y)", ylab = "Residuals (log Y)",
     main = "Residuals vs Fitted (log)")
abline(h = 0, lty = 2)

qqnorm(resid(fit_log), main = "Normal Q-Q (log residuals)")
qqline(resid(fit_log), lty = 2)
```

\newpage

## Step 4: Initial Model and Research Question Answers

With the Box-Cox transformation confirmed to have improved normality
diagnostics, the research questions can now be answered using ANOVAs on
the log-transformed model. Note that the response is now
log(ejection_fraction), so contributions are interpreted on the log
scale.

### Research question 1:

Does serum sodium contribute significantly to explaining ejection
fraction holding all other predictors constant?

### Research question 2:

Does serum sodium affect ejection rate differently in people with high
blood pressure compared to those with low blood pressure?

### Research question 3:

Does creatinine phosphokinase contribute significantly to explaining
ejection fraction holding all other predictors constant?

### Research question 4:

Does age contribute significantly to explaining ejection fraction
holding all other predictors constant?

### Research question 5:

Does serum creatinine contribute significantly to explaining ejection
fraction holding all other predictors constant?

```{r}
# Full Model (log-transformed)
summary(fit_log)

# Research Question 1:
reduced_model1 <- lm(
  log(ejection_fraction) ~ high_blood_pressure + serum_creatinine +
    creatinine_phosphokinase + age,
  data = df
)

full_model1 <- lm(
  log(ejection_fraction) ~ high_blood_pressure + serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)

anova(reduced_model1, full_model1)

# Research Question 2:
reduced_model2 <- lm(
  log(ejection_fraction) ~ high_blood_pressure + serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)

full_model2 <- lm(
  log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)

anova(reduced_model2, full_model2)

# Research Question 3:
reduced_model3 <- lm(
  log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
    serum_creatinine + age,
  data = df
)

full_model3 <- lm(
  log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)

anova(reduced_model3, full_model3)

# Research Question 4:
reduced_model4 <- lm(
  log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
    serum_creatinine + creatinine_phosphokinase,
  data = df
)

full_model4 <- lm(
  log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)

anova(reduced_model4, full_model4)

# Research Question 5:
reduced_model5 <- lm(
  log(ejection_fraction) ~ high_blood_pressure * serum_sodium + 
    creatinine_phosphokinase + age,
  data = df
)

full_model5 <- lm(
  log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
    serum_creatinine + creatinine_phosphokinase + age,
  data = df
)

anova(reduced_model5, full_model5)
```

\newpage

## Step 5: Fit the Final Model and Revisit Research Questions

After answering the research questions on the log-transformed model,
variable reduction is considered via criteria evaluation and
cross-validation:

```{r}
# Criteria Evaluation and Cross-Validation (Topic 19)

# Criteria Evaluation
y <- log(dat[["ejection_fraction"]])
dat$hbp_sodium <- with(dat, as.numeric(high_blood_pressure == "Yes") * serum_sodium)

fit_lm <- function(y, X) {
  df <- data.frame(y = y, X)
  lm(y ~ ., data = df)
}

press_stat <- function(model) {
  h <- hatvalues(model)
  e <- resid(model)
  sum((e/(1 - h))^2)
}

# Full model (for Cp baseline)
preds <- c("high_blood_pressure", "serum_sodium", "hbp_sodium",
           "serum_creatinine", "creatinine_phosphokinase", "age")
X_full <- dat[, preds, drop = FALSE]
m_full <- fit_lm(y, X_full)
SSE_full <- sum(resid(m_full)^2)
MSE_full <- SSE_full / df.residual(m_full)
n <- length(y)

# Enumerate all non-empty subsets and compute metrics
subsets <- unlist(lapply(1:length(preds), function(k) {
  combn(preds, k, simplify = FALSE)
}), recursive = FALSE)

results <- lapply(subsets, function(S) {
  Xs <- dat[, S, drop = FALSE]
  m <- fit_lm(y, Xs)
  SSE <- sum(resid(m)^2)
  p <- length(coef(m))      # includes intercept
  list(
    subset = paste(S, collapse = ", "),
    k_pred = length(S),     # number of predictors
    p_params = p,           # parameters incl. intercept
    R2 = summary(m)$r.squared,
    AdjR2 = summary(m)$adj.r.squared,
    Cp = SSE/MSE_full - (n - 2*p),
    AIC = AIC(m),
    BIC = BIC(m),
    PRESS = press_stat(m),
    SSE = SSE
  )
})

results_df <- do.call(rbind, lapply(results, as.data.frame))
results_df <- results_df[order(results_df$k_pred, results_df$AIC), ]

# Save results
write.csv(results_df, "ejection_fraction_model_selection_results.csv", row.names = FALSE)

# ---- Best subsets by criterion ----
best_R2    <- results_df[which.max(results_df$R2), ]
best_AdjR2 <- results_df[which.max(results_df$AdjR2), ]
gap_Cp     <- abs(results_df$Cp - results_df$p_params)
best_Cp    <- results_df[which.min(gap_Cp), ]
best_AIC   <- results_df[which.min(results_df$AIC), ]
best_BIC   <- results_df[which.min(results_df$BIC), ]
best_PRESS <- results_df[which.min(results_df$PRESS), ]

sink("ejection_fraction_best_by_criterion.txt")
cat("Response column: ejection_fraction\n")
cat("Predictors considered:", paste(preds, collapse=", "), "\n\n")
cat("Best subsets by criterion:\n")
print(list(
  R2 = list(k = best_R2$k_pred, subset = best_R2$subset),
  AdjR2 = list(k = best_AdjR2$k_pred, subset = best_AdjR2$subset),
  Cp = list(k = best_Cp$k_pred, subset = best_Cp$subset),
  AIC = list(k = best_AIC$k_pred, subset = best_AIC$subset),
  BIC = list(k = best_BIC$k_pred, subset = best_BIC$subset),
  PRESS = list(k = best_PRESS$k_pred, subset = best_PRESS$subset)
))
sink()

# Cross-Validation with 10-fold CV
set.seed(512)
ctrl <- trainControl(method = "cv", number = 10)

# Candidate models according to criteria:
m1 <- log(ejection_fraction) ~ serum_sodium
m4 <- log(ejection_fraction) ~ high_blood_pressure * serum_sodium + age
m6 <- log(ejection_fraction) ~ high_blood_pressure * serum_sodium +
  serum_creatinine + creatinine_phosphokinase + age

heart_failure <- read.csv("heart_failure_clinical_records_dataset 2.csv", stringsAsFactors = FALSE)

# 10-fold CV for OLS (method='lm')
fit1 <- train(m1, data = heart_failure, method = "lm", trControl = ctrl)
fit4 <- train(m4, data = heart_failure, method = "lm", trControl = ctrl)
fit6 <- train(m6, data = heart_failure, method = "lm", trControl = ctrl)

# RMSE Comparison
cat(sprintf("M1 RMSE: %.3f\n", fit1$results$RMSE[1]))
cat(sprintf("M4 RMSE: %.3f\n", fit4$results$RMSE[1]))
cat(sprintf("M6 RMSE: %.3f\n", fit6$results$RMSE[1]))
```

The three candidate models have identical performances (identical RMSE),
so for parsimony and interpretability, the final model is:
log(ejection_fraction) \~ serum_sodium

```{r}
summary(lm(log(ejection_fraction) ~ serum_sodium, data = heart_failure))
```

The fact that the final model only includes serum_sodium confirms the
results of all previous research questions as no other predictor was
found to be significant
